Model Recognition as Planning

### 2018-09-24 doc/icaps-19/model-recognition.tex
new stable version icaps19 paper. Ready for a second review

\maketitle
\begin{abstract} 
Given the partial observation of a plan execution and a set of possible planning models (that share the same state variables but define different action models), {\em model recognition} is the task of identifying which model in the set explains (produced) the given observation. The paper formalizes the {\em model recognition} task and proposes a novel method to assess the probability of a \strips\ model to produce a partial observation of a plan execution. This method, that we called {\em model recognition as planning}, is built on top of off-the-shelf classical planning algorithms, that are used to elicit the likelihood of the observations given a candidate model. {\em Model recognition as planning} is robust to missing intermediate states and actions in the observed plan execution. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different kinds of {\em finite state machines}. We show that {\em model recognition as planning} succeeds to identify the executed automata despite the internal machine state or actual applied transitions, are unobserved.
\end{abstract}

@@

This paper introduces the {\em model recognition} task where the target of the recognition process is not a goal but a {\em planning model}. Given a partial observation of a plan execution and a set of possible planning models (that share the same state variables but define different action models), {\em model recognition} is the task of identifying which model in the set has the highest probability of producing the given observation.

\begin{figure}
  \begin{scriptsize}    
  \begin{center}
  \begin{tikzpicture}[scale=.4]
          \begin{scope}
            \draw (0, 0) grid (5, 5);
            \node[anchor=center] at (3.5, 1.5) {R};
            \node[anchor=center] at (0.5, 0.5) {};
            \draw[thick,style=dotted] (0.5,0.5) -- (4.5,0.5);
            \draw[thick,style=dotted] (4.5,0.5) -- (4.5,1.5);
            \draw[thick,style=dotted,->] (4.5,1.5) -- (3.7,1.5);                        
          \end{scope}
        \end{tikzpicture}
\hspace*{1cm}        
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$q_0$};
	  \node[state] (B) [right of=A] {$q_1$};          
	  \path
(A) edge [bend left, align=center] node {inc-y,\\dec-y} (B)
(B) edge [bend left, align=center] node {inc-y,\\dec-y} (A);
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small Observation of the execution of a robot navigation plan in a $5\times 5$ grid and a two state automata for controlling that the robot only increments its x-coordinate at {\em even} rows (i.e. at state $q_0$).}
\label{fig:grid-example}
\end{figure}

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model the robot could increment its {\em x-coordinate} at {\em even} rows and decrement it at the {\em odd rows}. Apart from this particular navigation model, different models could be defined within the same state variables (e.g. altering the way {\small\tt (q0)} and {\small\tt (q1)} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution, like the one illustrated at Figure~\ref{fig:grid-example}, {\em model recognition} aims here to identify which navigation model produced that observation, despite key information is unobserved (e.g. the value of {\small\tt (q0)} and {\small\tt (q1)} or the particular applied actions).

@@

The paper introduces also {\em model recognition as planning}; a novel method to assess the probability of a given \strips\ model to produce an observed plan execution. Our method is built on top of off-the-shelf classical planning algorithms, that are used to elicit the likelihood of the observations given a candidate model. {\em Model recognition as planning} is robust to missing intermediate states and actions in the observed plan execution. We evaluate the effectiveness of {\em model recognition as planning} with a set of \strips\ models that represent different {\em finite state machines}. All of these {\em automatae} are defined within the same {\em alphabet} and same {\em machine states} but different {\em transition functions}. We show that {\em model recognition as planning} succeeds to identify the executed {\em automata} despite internal machine states or actual applied transitions are unobserved.

@@

  (:action dec-y-odd
    :parameters (?y1 ?y2)
    :precondition (and (ycoord ?y1) (next ?y2 ?y1) (q1))
    :effect (and (not (ycoord ?y1)) (ycoord ?y2)
                 (not (q1)) (q0)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Action model for a robot navigation in a $n\times n$ grid.}
\label{fig:model-example}
\end{figure}

\section{Background}
\label{sec:background}
This section formalizes the models for {\em classical planning} and for the {\em observations} of a plan execution.

@@

A {\em classical planning frame} is a pair $\tup{F,A}$, where $F$ is a set of fluents and $A$ is a set of actions whose semantics are specified with two functions: $\rho(s,a)$ that denotes whether an action $a\in A$ is {\em applicable} in a state $s$ and $\theta(s,a)$ that denotes the {\em successor state} that results of applying $a$ in $s$. In this work we specify action semantics with the \strips\ model. With this regard, an action $a\in A$ is defined with:
\begin{itemize}
\item $\pre(a)\in\mathcal{L}(F)$, the {\em preconditions} of $a$, is the set of literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\in\mathcal{L}(F)$, the {\em positive effects} of $a$, is the set of literals that are true after the application of action $a\in A$.
\item $\eff^-(a)\in\mathcal{L}(F)$, the {\em negative effects} of $a$, is the set of literals that are false after the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

@@

On the one hand, the initial state $I$ is fully observed while the observed states in $\tau$ may be partial, i.e. the value of certain fluents in the intermediate states may be omitted (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). On the other hand, the sequence of observed states $\tup{s_1^o, \ldots, s_m^o}$ in $\tau$ is the same sequence of states traversed by $\pi$ but certain states may also be omitted. Formally, $0\leq|s_i^o|$ for every $1\leq i\leq m$. This means that the transitions between two consecutive observed states in $\tau$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where $k\geq 1$ is unknown and unbound). Therefore we can conclude that having $\tau$ does not implies knowing the actual length of $\pi$.

\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\tau$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every $0\leq i\leq m$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

For instance, Figure~\ref{fig:grid-example} illustrates the $\Phi$-observation \{{\tt\scriptsize<(xcoord 1)(ycoord 1)>, <(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\} where $\Phi$ only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}. This means that, for each observed state, only the value of fluents {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} is known while the value of the fluents {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)} is unknown.

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\tau}$ where:
\begin{itemize}
\item $P=\tup{F,A,I,G}$ is a {\em classical planning problem} such that the semantics of the actions in $A$ is unknown because functions $\rho(s,a)$ and/or $\theta(s,a)$ are undefined.
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite and non-empty {\em set of models} for the actions in $A$ such that, each model in $\mathcal{M}\in M$, defines a different function pair $\tup{\rho,\theta}$.
\item $\tau$ is an {\em observation} of the execution of a solution plan $\pi$ for $P$.
\end{itemize}

The task of {\em model recognition} can be understood as a classification task where each class is represented with a different planning model and the observed plan execution is the example to classify. In this case, the planning model that is associated to a class is acting as a class prototype that the summarizes all the plan executions that could be synthesized with that model or in other words, all the examples that belong to that class.

In the {\em model recognition} task, hypotheses are about the possible action models $\mathcal{M}\in M$ while $\tau$ represents the given observation. The {\em discrete probability distribution} $P(\mathcal{M}|\tau)$ expresses, for each model $\mathcal{M}\in M$, its probability of producing the input observation. According to the {\em Bayes rule} this probability distribution is given by $P(\mathcal{M}|\tau)=\frac{P(\tau|\mathcal{M})P(\mathcal{M})}{P(\tau)}$ and can be estimated in these three steps:
\begin{enumerate}
\item Estimating the {\em a priori probabilities} (if they are not already given as input). $P(\tau)$, that measures how surprising is the given observation and $P(\mathcal{M})$, that expresses if one model is known to be a priori more likely than the others. 
\item Computing likelihoods $P(\tau|\mathcal{M})$.  Our approach is to estimate these likelihoods according to the {\em amount of edits} required by the model $\mathcal{M}$ to produce a plan $\pi$ such that:
\begin{enumerate}
\item $\pi$ solves $P$ and,
\item $\tau$ is a consistent observation of the execution of $\pi$ on the classical planning problem $P$. 
\end{enumerate}
The {\em edit distance} is a similarity metric that is traditionally computed over {\em strings} or {\em graphs} and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work this similarity metric refers to the edition of classical planning models.
\item Applying the {\em Bayes rule} to obtain the normalized posterior probabilities. The $P(\mathcal{M}|\tau)$ probabilities must sum 1 for all the $\mathcal{M}\in M$.
\end{enumerate}

The {\em naive Bayes classifier} assigns a model $\mathcal{M}\in M$ to the given observation $\tau$ wrt the following expression, $argmax_{\mathcal{M}\in M} P(\mathcal{M})\times P(\tau|\mathcal{M})$. The {\em solution} to the {\em model recognition} task is the model $\mathcal{M}\in M$ (or subset of models in $M$) that maximizes the previous expression.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (in other words, the corresponding $\rho$ and $\theta$ functions) are specified with \strips\ action schemas.

We start formalizing the \strips\ schema and define then the full space of possible \strips\ schema. Eventually, we introduce an {\em edit distance} for \strips\ schema which allows us to estimate the $P(\tau|\mathcal{M})$ likelihoods for planning models.

@@

A \strips\ action schema $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and the three list of predicates ($pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from the schema.

@@

Given a \strips\ action schema $\xi$, let us define an additional set of objects ($\Omega\cap\Omega_\xi=\emptyset$), that we denote as {\em variable names}, and that contains one variable name for each parameter in $pars(\xi)$, that is $\Omega_\xi=\{v_i\}_{i=1}^{|pars(\xi)|}$. Note that, for any of the actions schema defined in~\ref{fig:model-example}, $\Omega_\xi=\{v_1,v_2\}$.

The set of FOL interpretations of $\Psi$ (the {\em predicates} that shape the propositional state variables) over the corresponding $\Omega_\xi$ objects (the {\em variable names} for that schema), confines the elements that can appear in the $\xi$ \strips\ action schema, that is in each of its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists. We denote this set of FOL interpretations as ${\mathcal I}_{\Psi,\xi}$. For any of the actions schema defined in~\ref{fig:model-example} this set contains the same ten elements, ${\mathcal I}_{\Psi,\xi}=${\small\tt\{xcoord($v_1$), xcoord($v_2$), ycoord($v_1$), ycoord($v_2$), q0(), q1(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

Despite any element from ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of a schema $\xi$. The space of possible \strips\ schema for $\xi$ is bound further by a set of constraints ${\mathcal C}$ of the following three kinds: 
\begin{itemize}
\item {\em Syntactic constraints}. \strips\ constraints require negative effects appearing as preconditions, negative effects cannot be positive effects at the same time and also, positive effects cannot appear as preconditions. Formally, $\eff^-(a)\subseteq \pre(a)$, $\eff^-(a)\cap \eff^+(a)=\emptyset$ and $\pre(a)\cap \eff^+(a)=\emptyset$). Given exclusively these syntactic constraints, the size of the space of possible \strips\ schema is given by the expression, $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. For the navigation model of Figure~\ref{fig:model-example}, $2^{2\times 10}=1,048,576$ for every action schema.
\item {\em Domain-specific constraints}. One can also introduce domain-specific knowledge to more precisely bound the space of possible \strips\ schema for a particular domain. For instance, in a {\em robot navigation} model, like the one in Figure~\ref{fig:model-example}, predicates {\small\tt q0()} and {\small\tt q1()} are always exclusive so they cannot hold a the same time in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists. Further, {\small\tt next($v_1$,$v_1$)} and {\small\tt next($v_2$,$v_2$)} will not appear at any of these lists because the {\tt\small next} predicate is coding the {\em succesor} function for {\em natural numbers}. In this case, introducing this domain-specific constraints would reduce the size of the space of possible schema to $2^{2\times 7}=16384$, for every action schema.
\end{itemize}

Now we are ready to define what is a well-defined \strips\ action schema.
\begin{definition}[Well-defined \strips\ action schema]
Given a set of {\em predicates} $\Psi$, a list of action {\em parameters} $pars(\xi)$, and set of FOL constraints ${\mathcal C}$ we say that $\xi$ is a {\bf well-defined \strips\ action schema} iff its three lists $pre(\xi)\subseteq {\mathcal I}_{\Psi,\xi}$, $del(\xi)\subseteq{\mathcal I}_{\Psi,\xi}$ and $add(\xi)\subseteq{\mathcal I}_{\Psi,\xi}$ only contain elements in ${\mathcal I}_{\Psi,\xi}$ and they satisfy all the constraints in ${\mathcal C}$. 
\end{definition}

\subsection{Edit distances for the \strips\ planning model}
We define two edit \emph{operations} on a schema $\xi\in\mathcal{M}$ that belongs to a \strips\ action model $\mathcal{M}\in M$:
\begin{itemize}
\item {\em Deletion}. An element is removed from any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schema $\xi\in\mathcal{M}$ such that the resulting schema is a {\em well-defined} \strips\ action schema.
\item {\em Insertion}. An element in ${\mathcal I}_{\Psi,\xi}$ is added to any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schema $\xi\in\mathcal{M}$ s.t. the resulting schema is {\em well-defined}.
\end{itemize}

@@

The observation of plan executions, generated with an action model $\mathcal{M}$, is {\em semantic knowledge} that constrain further the space of possible schema $\xi\in \mathcal{M}$. In this case, we talk about {\em observation constraints} that can also be added to the $\mathcal{C}$ set. In addition, this new kind of model constraints allow us to define an edit distance to asses the matching of a \strips\ model with respect to an observation of a plan execution.

\begin{definition}[Observation edit distance]
  Given $\tau$, an observation of the execution of a plan for solving $P$ and a \strips\ action model $\mathcal{M}$, all defined within the same set of predicates $\Psi$. The {\bf observation edit distance}, $\delta(\mathcal{M},\tau)$, is the minimal edit distance from $\mathcal{M}$ to any {\em comparable} model $\mathcal{M}'$ s.t. $\mathcal{M}'$ produces a plan $\pi^*_\tau$ optimal for $P$ and compliant with $\tau$; \[\delta(\mathcal{M},\tau)=\min_{\forall \mathcal{M}' \rightarrow \tau} \delta(\mathcal{M},\mathcal{M}')\]
\end{definition}

@@

\section{Model recognition with classical planning}
This section shows that, for \strips\ planning models, the $P(\tau|\mathcal{M})$ likelihood can be estimated with a classical planning compilation. In more detail, when the dynamics of the actions $A\in P$ are specified with \strips\ action schemas, then $\delta(\mathcal{M},\tau)$ can be computed with a compilation of a classical planning with conditional effects.

The compilation is an extension of the classical planning compilation for the learning of \strips\ planning models~\cite{aineto2018learning}. The intuition behind this compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the action model $\mathcal{M}$ to build $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that modifies the preconditions and effects of the action schemes in $\mathcal{M}$ using to the two {\em edit operations} defined above, {\em deletion} and {\em insertion}. 
\item {\bf Validates the edited model $\mathcal{M}'$}. The solution plan continues with a postfix that:
\begin{enumerate}
\item Induces an optimal solution plan $\pi^*_\tau$ for the original classical planning problem $P$.
\item Validates that $\tau$ is an observation of the execution of $\pi^*_\tau$ on the classical planning problem $P$.
\end{enumerate}
\end{enumerate}

>>>>

\maketitle
\begin{abstract} 
Given the partial observation of a plan execution and a set of possible planning models (that share the same state variables but define different action models to update these variables), {\em model recognition} is the task of identifying which model in the set explains (produced) the given observation. The paper formalizes the {\em model recognition} task and proposes a novel method to estimate the probability of a \strips\ model to produce a partial observation of a plan execution. This method, that we called {\em model recognition as planning}, is built on top of off-the-shelf classical planning algorithms and elicit the likelihood of the observation of a plan execution given a candidate model. {\em Model recognition as planning} is robust to missing intermediate states and actions in the observed plan execution. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different kinds of {\em finite state machines}. We show that {\em model recognition as planning} succeeds to identify the executed automata despite the internal machine state or actual applied transitions, are unobserved.
\end{abstract}

@@

This paper introduces the {\em model recognition} task where the object to recognize is not a goal but a {\em planning model}. Given a partial observation of a plan execution and a set of possible planning models (that share the same state variables but define different action models to update these variables), {\em model recognition} is the task of identifying which model in the set has the highest probability of explaining (producing) the given observation.

\begin{figure}
  \begin{scriptsize}    
  \begin{center}
  \begin{tikzpicture}[scale=.4]
          \begin{scope}
            \draw (0, 0) grid (5, 5);
            \node[anchor=center] at (3.5, 1.5) {R};
            \node[anchor=center] at (0.5, 0.5) {};
            \draw[thick,style=dotted] (0.5,0.5) -- (4.5,0.5);
            \draw[thick,style=dotted] (4.5,0.5) -- (4.5,1.5);
            \draw[thick,style=dotted,->] (4.5,1.5) -- (3.7,1.5);                        
          \end{scope}
        \end{tikzpicture}
\hspace*{1cm}        
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$q_0$};
	  \node[state] (B) [right of=A] {$q_1$};          
	  \path
(A) edge [bend left, align=center] node {inc-y,\\dec-y} (B)
(B) edge [bend left, align=center] node {inc-y,\\dec-y} (A);
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small Observation of the execution of a robot navigation plan in a $5\times 5$ grid and a 2-state automata for controlling that the robot only increments its x-coordinate at {\em even} rows (i.e. when $q_0$ holds).}
\label{fig:grid-example}
\end{figure}

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model the robot could increment its {\em x-coordinate} at {\em even} rows (when {\tt\small $q0$} holds) and decrement it at the {\em odd rows} (when {\tt\small $q1$} holds). Apart from this particular navigation model, different action models could be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution, like the one illustrated at Figure~\ref{fig:grid-example}, {\em model recognition} aims here to identify which navigation model produced that observation, despite key information is unobserved (e.g. the particular applied actions or the value of {\small\tt $q0$} and {\small\tt $q1$}).

@@

The paper introduces also {\em model recognition as planning}; a novel method to estimate the probability of a given \strips\ model to produce an observed plan execution. Our method is built on top of off-the-shelf classical planning algorithms, that are used to elicit the likelihood of the observations given a candidate model. {\em Model recognition as planning} is robust to missing intermediate states and actions in the observed plan execution.

We evaluate the effectiveness of {\em model recognition as planning} with a set of \strips\ models that represent different {\em finite state machines}. All of these {\em automatae} are defined within the same input {\em alphabet} and same {\em machine states} but different {\em transition functions}. We show that {\em model recognition as planning} succeeds to identify the executed {\em automata} despite internal machine states or actual applied transitions are unobserved.

@@

  (:action dec-y-odd
    :parameters (?y1 ?y2)
    :precondition (and (ycoord ?y1) (next ?y2 ?y1) (q1))
    :effect (and (not (ycoord ?y1)) (ycoord ?y2)
                 (not (q1)) (q0)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Example of a \strips\ action model (codded in PDDL) for robot navigation in a $n\times n$ grid.}
\label{fig:model-example}
\end{figure}

\section{Background}
\label{sec:background}
This section formalizes the models for {\em classical planning} and for the {\em observation} of a plan execution.

@@

A {\em classical planning frame} is a pair $\tup{F,A}$, where $F$ is a set of fluents and $A$ is a set of actions whose semantics are specified with two functions: $\rho(s,a)$ that denotes whether an action $a\in A$ is {\em applicable} in a state $s$ and $\theta(s,a)$ that denotes the {\em successor state} that results of applying $a$ in $s$. In this work we specify action semantics (the $\rho$ and $\theta$ functions) with the \strips\ model. With this regard, an action $a\in A$ is defined with:
\begin{itemize}
\item $\pre(a)\in\mathcal{L}(F)$, the {\em preconditions} of $a$, is the set of literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\in\mathcal{L}(F)$, the {\em positive effects} of $a$, is the set of literals that are true after the application of action $a\in A$.
\item $\eff^-(a)\in\mathcal{L}(F)$, the {\em negative effects} of $a$, is the set of literals that are false after the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

@@

On the one hand, the initial state $I$ is fully observed while the observed states in $\tau$ may be partial, i.e. the value of certain fluents in the intermediate states may be omitted (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). On the other hand, the sequence of observed states $\tup{s_1^o, \ldots, s_m^o}$ in $\tau$ is the same sequence of states traversed by $\pi$ but certain states may also be omitted. Formally, $0\leq|s_i^o|$ for every $1\leq i\leq m$. This means that the transitions between two consecutive observed states in $\tau$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown and unbound). Therefore we can conclude that having $\tau$ does not implies knowing the actual length of $\pi$.

\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\tau$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 0\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

For instance, Figure~\ref{fig:grid-example} illustrates the six-state $\Phi$-observation \{{\tt\footnotesize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\} where $\Phi$ only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}. This means that, for each observed state, only the value of fluents {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} is known while the value of the remaining fluents {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)} is unknown.

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\tau}$ where:
\begin{itemize}
\item $P=\tup{F,A,I,G}$ is a {\em classical planning problem} s.t. the semantics of actions $a\in A$ is unknown because the corresponding functions, $\rho(s,a)$ and/or $\theta(s,a)$, are undefined.
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite and non-empty {\em set of models} for the actions in $A$ s.t. each model in $\mathcal{M}\in M$, defines a different function pair $\tup{\rho,\theta}$.
\item $\tau$ is an {\em observation} of the execution of a solution plan $\pi$ for $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model and the observed plan execution is the single example to classify. In this case, the planning model that is associated to a class is acting as the {\em class prototype} and summarizes all the plan executions that could be synthesized with that model (all the examples that belong to that class).

In {\em model recognition}, hypotheses are then about the possible action models $\mathcal{M}\in M$ while the observation $\tau$ of a plan execution represents the input observation. The {\em naive Bayes classifier} assigns a model $\mathcal{M}\in M$ to the given observation $\tau$ wrt the following expression, $argmax_{\mathcal{M}\in M} P(\mathcal{M})\times P(\tau|\mathcal{M})$. The {\em solution} to the {\em model recognition} task is the model $\mathcal{M}\in M$ (or subset of models in $M$) that maximizes the previous expression.

The $P(\mathcal{M})$ probability expresses if one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that {\em a priori} all models are equiprobable. Our approach for {\em model recognition} is then to estimate the $P(\tau|\mathcal{M})$ likelihood according to the {\em amount of edits} required by the model $\mathcal{M}$ to produce a plan $\pi$ such that:
\begin{enumerate}
\item $\pi$ solves $P$ and,
\item $\tau$ is a {\em consistent} observation of the execution of $\pi$ on the classical planning problem $P$. 
\end{enumerate}
The {\em edit distance} is a similarity metric that is traditionally computed over {\em strings} or {\em graphs} and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work we show how to compute this similarity metric referred to the edition of classical planning models.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (i.e. $\rho$ and $\theta$ functions) are specified with \strips\ action schemas.

We start formalizing the \strips\ schema and define the full space of possible \strips\ schema. Eventually, we introduce an {\em edit distance} for \strips\ schema to estimate the $P(\tau|\mathcal{M})$ likelihoods for classical planning models.

@@

A \strips\ action schema $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and three list of predicates ($pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

@@

Given a \strips\ action schema $\xi$, let us define an additional set of objects ($\Omega\cap\Omega_\xi=\emptyset$), that we denote as {\em variable names}, and that contains one variable name for each parameter in $pars(\xi)$, that is $\Omega_\xi=\{v_i\}_{i=1}^{|pars(\xi)|}$. Note that, for any of the six schema defined in Figure~\ref{fig:model-example}, $\Omega_\xi=\{v_1,v_2\}$.

Given a \strips\ action schema $\xi$ and a set of {\em predicates} $\Psi$ that shape the propositional state variables. The set of FOL interpretations of $\Psi$ over the corresponding $\Omega_\xi$ objects (the {\em variable names} for schema $\xi$), confines the elements that can appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists. We denote this set of FOL interpretations as ${\mathcal I}_{\Psi,\xi}$. For any of the six schema defined in Figure~\ref{fig:model-example} the ${\mathcal I}_{\Psi,\xi}$ set contains the same ten elements, ${\mathcal I}_{\Psi,\xi}=${\small\tt\{xcoord($v_1$), xcoord($v_2$), ycoord($v_1$), ycoord($v_2$), q0(), q1(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

Despite any element from ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of a schema $\xi$. The space of possible \strips\ schema is bound further by a set of constraints ${\mathcal C}$ of the following kinds: 
\begin{itemize}
\item {\em Syntactic constraints}. \strips\ constraints require negative effects appearing as preconditions, negative effects cannot be positive effects at the same time and also, positive effects cannot appear as preconditions. Formally, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$). Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schema is given by the expression, $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. For the navigation model of Figure~\ref{fig:model-example}, $2^{2\times 10}=1,048,576$ for every action schema.
\item {\em Domain-specific constraints}. One can also introduce domain-specific knowledge to more precisely bound the space of possible \strips\ schema for a particular domain. For instance, in a {\em robot navigation} model, like the one in Figure~\ref{fig:model-example}, predicates {\small\tt q0()} and {\small\tt q1()} are exclusive so they cannot hold a the same time in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists. Further, {\small\tt next($v_1$,$v_1$)} and {\small\tt next($v_2$,$v_2$)} will not appear at any of these lists because the {\tt\small next} predicate is coding the {\em succesor} function for {\em natural numbers}. In this case, introducing these domain-specific constraints would reduce the size of the space of possible schema to $2^{2\times 7}=16,384$ for every action schema.
\end{itemize}

Now we are ready to define what is a well-defined \strips\ action schema.
\begin{definition}[Well-defined \strips\ action schema]
Given a set of {\em predicates} $\Psi$, a list of action {\em parameters} $pars(\xi)$, and set of FOL constraints ${\mathcal C}$ we say that $\xi$ is a {\bf well-defined \strips\ action schema} iff its three lists $pre(\xi)\subseteq {\mathcal I}_{\Psi,\xi}$, $del(\xi)\subseteq{\mathcal I}_{\Psi,\xi}$ and $add(\xi)\subseteq{\mathcal I}_{\Psi,\xi}$ only contain elements in ${\mathcal I}_{\Psi,\xi}$ and they satisfy all the constraints in ${\mathcal C}$. 
\end{definition}
We say that an action model is well-defined if all its \strips\ action schema are well-defined.

\subsection{Edit distances for \strips\ action models}
We define two edit \emph{operations} on a schema $\xi\in\mathcal{M}$ that belongs to a \strips\ action model $\mathcal{M}\in M$:
\begin{itemize}
\item {\em Deletion}. An element is removed from any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schema $\xi\in\mathcal{M}$ such that the resulting schema is a {\em well-defined} \strips\ action schema.
\item {\em Insertion}. An element in ${\mathcal I}_{\Psi,\xi}$ is added to any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schema $\xi\in\mathcal{M}$ s.t. the resulting schema is {\em well-defined}.
\end{itemize}

@@

The observation of a plan execution, generated with an action model $\mathcal{M}$, reflects {\em semantic knowledge} that constrain further the space of the possible schema $\xi\in \mathcal{M}$. In this case, we talk about {\em observation constraints} that can also be added to the $\mathcal{C}$ set. In addition, this new kind of model constraints allow us to define an edit distance to asses the matching of a \strips\ model with respect to an observation of a plan execution.

\begin{definition}[Observation edit distance]
  Given $\tau$, an observation of the execution of a plan for solving $P$ and a \strips\ action model $\mathcal{M}$, all defined within the same set of predicates $\Psi$. The {\bf observation edit distance}, $\delta(\mathcal{M},\tau)$, is the minimal edit distance from $\mathcal{M}$ to any {\em comparable} and well-defined model $\mathcal{M}'$ s.t. $\mathcal{M}'$ produces a plan $\pi_\tau$ for $P$ {\em consistent} with $\tau$; \[\delta(\mathcal{M},\tau)=\min_{\forall \mathcal{M}' \rightarrow \tau} \delta(\mathcal{M},\mathcal{M}')\]
\end{definition}

@@

\section{Model recognition with classical planning}
This section shows that, for \strips\ planning models, $\delta(\mathcal{M},\tau)$ can be computed with a compilation of a classical planning with conditional effects (and hence the $P(\tau|\mathcal{M})$ likelihood can be estimated).

The compilation is an extension of the classical planning compilation for the learning of \strips\ planning models~\cite{aineto2018learning}. The intuition behind this compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the action model $\mathcal{M}$ to build $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that modifies the preconditions and effects of the action schemes in $\mathcal{M}$ using to the two {\em edit operations} defined above, {\em deletion} and {\em insertion}. 
\item {\bf Validates the edited model $\mathcal{M}'$}. The solution plan continues with a postfix that:
\begin{enumerate}
\item Induces a solution plan $\pi_\tau$ for the original classical planning problem $P$.
\item Validates that $\tau$ is an observation of the execution of $\pi_\tau$ on the classical planning problem $P$.
\end{enumerate}
\end{enumerate}
